taxize - taxonomic search and retrieval in R

### 2013-05-29 inst/doc/taxize_withcode_app.tex
first scratch of appB

Suppose we have fuzzy coded trait table with 2 traits with 3 respectively 2 modalities:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
(traits <- \hlfunctioncall{read.table}(header = TRUE, sep = \hlstring{";"}, text = \hlstring{"taxon;T1M1;T1M2;T1M3;T2M1;T2M2\textbackslash{}nGammarus sp.;0;0;3;1;3\textbackslash{}nPotamopyrgus antipodarum;1;0;3;1;3\textbackslash{}nCoenagrion sp.;3;0;1;3;1\textbackslash{}nEnallagma cyathigerum;0;3;1;0;3\textbackslash{}nErythromma sp.;0;0;3;3;1\textbackslash{}n"}))
\end{alltt}
\begin{verbatim}
                     taxon T1M1 T1M2 T1M3 T2M1 T2M2
1             Gammarus sp.    0    0    3    1    3
2 Potamopyrgus antipodarum    1    0    3    1    3
3           Coenagrion sp.    3    0    1    3    1
4    Enallagma cyathigerum    0    3    1    0    3
5           Erythromma sp.    0    0    3    3    1
\end{verbatim}
\end{kframe}
\end{knitrout}

And want to match this to a table with abundances:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
(abundances <- \hlfunctioncall{read.table}(header = TRUE, sep = \hlstring{";"}, text = \hlstring{"taxon;abundance\textbackslash{}nGammarus roeseli;5\textbackslash{}nGammarus roeseli;6\textbackslash{}nGammarus tigrinus;7\textbackslash{}nGammarus tigrinus;8\textbackslash{}nCoenagrionidae;10\textbackslash{}nCoenagrionidae;6\textbackslash{}nPotamopyrgus antipodarum;10\textbackslash{}n"}))
\end{alltt}
\begin{verbatim}
                     taxon abundance
1         Gammarus roeseli         5
2         Gammarus roeseli         6
3        Gammarus tigrinus         7
4        Gammarus tigrinus         8
5           Coenagrionidae        10
6           Coenagrionidae         6
7 Potamopyrgus antipodarum        10
\end{verbatim}
\end{kframe}
\end{knitrout}

>>>>

Suppose we have fuzzy coded trait table with 2 traits with 3 respectively 2 modalities:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
(traits <- \hlfunctioncall{read.table}(header = TRUE, sep = \hlstring{";"}, stringsAsFactors = FALSE, text = \hlstring{"taxon;T1M1;T1M2;T1M3;T2M1;T2M2\textbackslash{}nGammarus sp.;0;0;3;1;3\textbackslash{}nPotamopyrgus antipodarum;1;0;3;1;3\textbackslash{}nCoenagrion sp.;3;0;1;3;1\textbackslash{}nEnallagma cyathigerum;0;3;1;0;3\textbackslash{}nErythromma sp.;0;0;3;3;1\textbackslash{}nBaetis sp.;0;0;0;0;0\textbackslash{}n"}))
\end{alltt}
\begin{verbatim}
                     taxon T1M1 T1M2 T1M3 T2M1 T2M2
1             Gammarus sp.    0    0    3    1    3
2 Potamopyrgus antipodarum    1    0    3    1    3
3           Coenagrion sp.    3    0    1    3    1
4    Enallagma cyathigerum    0    3    1    0    3
5           Erythromma sp.    0    0    3    3    1
6               Baetis sp.    0    0    0    0    0
\end{verbatim}
\end{kframe}
\end{knitrout}

And want to match this to a table with abundances:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
(abundances <- \hlfunctioncall{read.table}(header = TRUE, sep = \hlstring{";"}, stringsAsFactors = FALSE, 
    text = \hlstring{"taxon;abundance;sample\textbackslash{}nGammarus roeseli;5;1\textbackslash{}nGammarus roeseli;6;2\textbackslash{}nGammarus tigrinus;7;1\textbackslash{}nGammarus tigrinus;8;2\textbackslash{}nCoenagrionidae;10;1\textbackslash{}nCoenagrionidae;6;2\textbackslash{}nPotamopyrgus antipodarum;10;1\textbackslash{}nxxxxx;10;2\textbackslash{}n"}))
\end{alltt}
\begin{verbatim}
                     taxon abundance sample
1         Gammarus roeseli         5      1
2         Gammarus roeseli         6      2
3        Gammarus tigrinus         7      1
4        Gammarus tigrinus         8      2
5           Coenagrionidae        10      1
6           Coenagrionidae         6      2
7 Potamopyrgus antipodarum        10      1
8                    xxxxx        10      2
\end{verbatim}
\end{kframe}
\end{knitrout}

First we do some basic data-cleaning and create a looktable, that will link taxa in trait table with taxa in the abundance table.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# first we remove ' sp.' from out trait table:}
traits$taxon_cleaned <- \hlfunctioncall{tolower}(\hlfunctioncall{gsub}(\hlstring{" sp."}, \hlstring{""}, traits$taxon))

\hlcomment{# since abundance tables can be very long with repeating taxa, we look}
\hlcomment{# only at unique taxon names This will be a lookup-table linking taxon}
\hlcomment{# names between both tables}
lookup <- \hlfunctioncall{data.frame}(taxon = \hlfunctioncall{tolower}(\hlfunctioncall{unique}(abundances$taxon)), stringsAsFactors = FALSE)
\end{alltt}
\end{kframe}
\end{knitrout}

The we query the taxonomic hierarchy for both tables, since this is the backbone of this procedure:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{require}(taxize)
traits_classi <- \hlfunctioncall{classification}(\hlfunctioncall{get_uid}(traits$taxon_cleaned))
lookup_classi <- \hlfunctioncall{classification}(\hlfunctioncall{get_uid}(lookup$taxon))
\end{alltt}
\end{kframe}
\end{knitrout}

First we look if we get any direct matches between taxon names:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# first search for direct matches}
direct <- \hlfunctioncall{match}(lookup$taxon, traits$taxon_cleaned)
\hlcomment{# and add the matched name to our lookup table}
lookup$traits <- \hlfunctioncall{tolower}(traits$taxon[direct])
lookup$type <- \hlfunctioncall{ifelse}(!\hlfunctioncall{is.na}(direct), \hlstring{"direct"}, NA)
lookup
\end{alltt}
\begin{verbatim}
                     taxon                   traits   type
1         gammarus roeseli                     <NA>   <NA>
2        gammarus tigrinus                     <NA>   <NA>
3           coenagrionidae                     <NA>   <NA>
4 potamopyrgus antipodarum potamopyrgus antipodarum direct
5                    xxxxx                     <NA>   <NA>
\end{verbatim}
\end{kframe}
\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# look for cases where taxonomic resolution in abundance data is higher}
\hlcomment{# than in trait data: here we take the trait-values for the lower}
\hlcomment{# resolution}
\hlfunctioncall{for} (i in \hlfunctioncall{which}(\hlfunctioncall{is.na}(lookup$traits))) \{
    \hlfunctioncall{if} (\hlfunctioncall{is.data.frame}(lookup_classi[[i]])) \{
        matches <- \hlfunctioncall{tolower}(lookup_classi[[i]]$ScientificName) %in% traits$taxon_cleaned
        \hlfunctioncall{if} (\hlfunctioncall{any}(matches)) \{
            lookup$traits[i] <- \hlfunctioncall{tolower}(lookup_classi[[i]]$ScientificName[matches])
            lookup$type[i] <- lookup_classi[[i]]$Rank[matches]
        \}
    \}
\}
lookup
\end{alltt}
\begin{verbatim}
                     taxon                   traits   type
1         gammarus roeseli                 gammarus  genus
2        gammarus tigrinus                 gammarus  genus
3           coenagrionidae                     <NA>   <NA>
4 potamopyrgus antipodarum potamopyrgus antipodarum direct
5                    xxxxx                     <NA>   <NA>
\end{verbatim}
\begin{alltt}

\hlcomment{# look for cases taxonomic resolution in abundance data is lower than in}
\hlcomment{# trait data, here we need to aggregate the trait-values (eg. median value}
\hlcomment{# for modality)}

\hlfunctioncall{for} (i in \hlfunctioncall{which}(\hlfunctioncall{is.na}(lookup$traits))) \{
\hlcomment{    # find matches}
    agg <- \hlfunctioncall{sapply}(traits_classi, \hlfunctioncall{function}(x) \hlfunctioncall{any}(\hlfunctioncall{tolower}(x$ScientificName) %in% 
        lookup$taxon[i]))
    \hlfunctioncall{if} (\hlfunctioncall{sum}(agg) > 1) \{
\hlcomment{        # add aggregated trait to trait table}
        traits <- \hlfunctioncall{rbind}(traits, \hlfunctioncall{c}(\hlfunctioncall{paste0}(lookup$taxon[i], \hlstring{"-aggregated"}), \hlfunctioncall{apply}(traits[agg, 
            2:6], 2, median), \hlfunctioncall{paste0}(lookup$taxon[i], \hlstring{"-aggregated"})))
\hlcomment{        # fill lookup table}
        lookup$traits[i] <- \hlfunctioncall{paste0}(lookup$taxon[i], \hlstring{"-aggregated"})
        lookup$type[i] <- \hlstring{"aggregated"}
    \}
\}
lookup
\end{alltt}
\begin{verbatim}
                     taxon                    traits       type
1         gammarus roeseli                  gammarus      genus
2        gammarus tigrinus                  gammarus      genus
3           coenagrionidae coenagrionidae-aggregated aggregated
4 potamopyrgus antipodarum  potamopyrgus antipodarum     direct
5                    xxxxx                      <NA>       <NA>
\end{verbatim}
\begin{alltt}

abundances$traits <- \hlfunctioncall{tolower}(lookup$traits[\hlfunctioncall{match}(\hlfunctioncall{tolower}(abundances$taxon), 
    lookup$taxon)])
abundances
\end{alltt}
\begin{verbatim}
                     taxon abundance sample                    traits
1         Gammarus roeseli         5      1                  gammarus
2         Gammarus roeseli         6      2                  gammarus
3        Gammarus tigrinus         7      1                  gammarus
4        Gammarus tigrinus         8      2                  gammarus
5           Coenagrionidae        10      1 coenagrionidae-aggregated
6           Coenagrionidae         6      2 coenagrionidae-aggregated
7 Potamopyrgus antipodarum        10      1  potamopyrgus antipodarum
8                    xxxxx        10      2                      <NA>
\end{verbatim}
\begin{alltt}

traits[\hlfunctioncall{match}(abundances$traits, traits$taxon_cleaned), 2:6]
\end{alltt}
\begin{verbatim}
    T1M1 T1M2 T1M3 T2M1 T2M2
1      0    0    3    1    3
1.1    0    0    3    1    3
1.2    0    0    3    1    3
1.3    0    0    3    1    3
7      0    0    1    3    1
7.1    0    0    1    3    1
2      1    0    3    1    3
NA  <NA> <NA> <NA> <NA> <NA>
\end{verbatim}
\end{kframe}
\end{knitrout}

### 2013-05-29 inst/doc/taxize_withcode_appB.Rnw
first scratch of appB

Suppose we have fuzzy coded trait table with 2 traits with 3 respectively 2 modalities:
<<>>=
(traits <- read.table(header = TRUE, sep = ';', text = 'taxon;T1M1;T1M2;T1M3;T2M1;T2M2
Gammarus sp.;0;0;3;1;3
Potamopyrgus antipodarum;1;0;3;1;3
Coenagrion sp.;3;0;1;3;1
Enallagma cyathigerum;0;3;1;0;3
Erythromma sp.;0;0;3;3;1
'))
@

And want to match this to a table with abundances:
<<>>=
(abundances <- read.table(header = TRUE, sep = ';', text = 'taxon;abundance
Gammarus roeseli;5
Gammarus roeseli;6
Gammarus tigrinus;7
Gammarus tigrinus;8
Coenagrionidae;10
Coenagrionidae;6
Potamopyrgus antipodarum;10
'))
@

>>>>

Suppose we have fuzzy coded trait table with 2 traits with 3 respectively 2 modalities:
<<traitdata>>=
(traits <- read.table(header = TRUE, sep = ';', stringsAsFactors=FALSE, text = 'taxon;T1M1;T1M2;T1M3;T2M1;T2M2
Gammarus sp.;0;0;3;1;3
Potamopyrgus antipodarum;1;0;3;1;3
Coenagrion sp.;3;0;1;3;1
Enallagma cyathigerum;0;3;1;0;3
Erythromma sp.;0;0;3;3;1
Baetis sp.;0;0;0;0;0
'))
@

And want to match this to a table with abundances:
<<abundancedata>>=
(abundances <- read.table(header = TRUE, sep = ';', stringsAsFactors=FALSE, text = 'taxon;abundance;sample
Gammarus roeseli;5;1
Gammarus roeseli;6;2
Gammarus tigrinus;7;1
Gammarus tigrinus;8;2
Coenagrionidae;10;1
Coenagrionidae;6;2
Potamopyrgus antipodarum;10;1
xxxxx;10;2
'))
@

First we do some basic data-cleaning and create a looktable, that will link taxa in trait table with taxa in the abundance table.
<<cleaning>>=
# first we remove ' sp.' from out trait table:
traits$taxon_cleaned <- tolower(gsub(' sp.', '', traits$taxon))

# since abundance tables can be very long with repeating taxa, we look only at unique
# taxon names 
# This will be a lookup-table linking taxon names between both tables
lookup <- data.frame(taxon = tolower(unique(abundances$taxon)), stringsAsFactors=FALSE)
@

The we query the taxonomic hierarchy for both tables, since this is the backbone of this procedure:
<<get_classification, cache=TRUE>>=
require(taxize)
traits_classi <- classification(get_uid(traits$taxon_cleaned))
lookup_classi <- classification(get_uid(lookup$taxon))
@

First we look if we get any direct matches between taxon names:
<<direct_matches>>=
# first search for direct matches
direct <- match(lookup$taxon, traits$taxon_cleaned)
# and add the matched name to our lookup table
lookup$traits <- tolower(traits$taxon[direct])
lookup$type <- ifelse(!is.na(direct), "direct", NA)
lookup
@

<<>>=
# look for cases where taxonomic resolution in abundance data is higher than in trait data:
# here we take the trait-values for the lower resolution
for(i in which(is.na(lookup$traits))){
  if(is.data.frame(lookup_classi[[i]])){
    matches <- tolower(lookup_classi[[i]]$ScientificName) %in% traits$taxon_cleaned
    if(any(matches)){
      lookup$traits[i] <- tolower(lookup_classi[[i]]$ScientificName[matches])
      lookup$type[i] <- lookup_classi[[i]]$Rank[matches]
    }
  }
}
lookup

# look for cases taxonomic resolution in abundance data is lower than in trait data,
# here we need to aggregate the trait-values (eg. median value for modality)

for(i in which(is.na(lookup$traits))){
  # find matches
  agg <- sapply(traits_classi, function(x) any(tolower(x$ScientificName) %in% lookup$taxon[i]))
  if(sum(agg) > 1){
    # add aggregated trait to trait table
    traits <- rbind(traits, c(paste0(lookup$taxon[i], '-aggregated'), apply(traits[agg, 2:6], 2, median), paste0(lookup$taxon[i], '-aggregated')))
    # fill lookup table
    lookup$traits[i] <- paste0(lookup$taxon[i], '-aggregated')
    lookup$type[i] <- 'aggregated'
  }
}
lookup

abundances$traits <- tolower(lookup$traits[match(tolower(abundances$taxon), lookup$taxon)])
abundances

traits[match(abundances$traits, traits$taxon_cleaned), 2:6]
@
